### 2017-03-24 chapters/introduction/patient.tex
Update thesis to latest version

\section{Patient Specific Model Generation}

In this section we will describe

@@

The geometry of a patient's heart can be aquired using medical
imaging techniques, such as echocardiography (ultracound), magnetic
resonance imaging (MRI) or computed tomography (CT). Each modality off
advantages and disadvatages over the other. For example, MRI provides
high quality images, uses zero radiation, but is exspensive and lacks
temporal resolution. CT can more accuratley reconstruct the 3D image
in contrast to MRI, in which 2D slices needs to be glued together to
form a 3D surface. However, CT exposes the patient to radiation which
increases the chance of developing cancer. Finally echocardiography is
easy to use, cheap, harmless, and  has good temporal resolution, but
is clearly inferior when it comes to image quality.

The main modality used in this thesis is 3D echocardiography

@@

\subsection{Continuum description of the heart}

\begin{itemize}
  \item Reference to mehcanics book
  \item Introduce basic notation (stress, strain, deformation)
\end{itemize}

We represent the heart as a continuum body  embedded in
$\mathbb{R}^3$. Let $\Xvec$ and  $\xvec$ denote the corresponding
coordinate in the reference and current configuration respectively.
The motion of a point $\Xvec$ in the reference configuration to the
point $\xvec$ in the current configuration  may be described by a
deformation map  $\varphi :  \mathcal{B}_0  \rightarrow \mathcal{B}$.
The deformation gradient associated with the motion $\xvec =
\varphi(\Xvec)$ is a rank-2 tensor given by  
\begin{align}
\F(\Xvec) = \frac{\partial \varphi}{\partial \Xvec}, 
\end{align}
The deformed and reference configuration are related via the
displacement field $\uvec$ by: 
\begin{align}  
\uvec = \xvec-\Xvec = \chi( \Xvec, t) -\Xvec,
\end{align} 
Further, we let $J = \det \F$ be the determinant of the deformation gradient 
and $\C = \F^T\F$ the right Cauchy-Green deformation tensor.\cite{holzapfel2000nonlinear}

\subsection{Passive behavior of the myocardium}

\begin{itemize}
  \item Hyperelasticiy, fung type-refer to relevant papers
  \item Incompressibility - Saddle point problem
  \item Finding the unloaded configuration
\end{itemize}

\subsection{Active behavior of the myocardium}

\begin{itemize}
  \item Active stress and strain (some mathematical aspects perhaps)
\end{itemize}

\subsection{Adjoint-based Data Assimilation}

\begin{itemize}
  \item Examples of data you want to constrain
  \item Forming of mismatch functional and problem formulation as
    PDE-constrained optimization.
  \item The adjoint approach. comparison with other methods.
  \item Control parameters
  \item Reference to pulse{\_}adjoint and pulse{\_}adjoint{\_}post
    repositories
\end{itemize}

@@

\begin{equation}
  \begin{aligned}
    \label{eq:opt_matparam}
    & \underset{m}{\text{minimize}}
    & &  I(\state, m) \\
    & \text{subject to}
    & & \delta \Pi(\state, \mvec) = 0, 
  \end{aligned}
\end{equation}

where $I(\state, m): \mathbb{W} \times \mathbb{Q} \mapsto
\mathbb{R}$ for some state space $\mathbb{W}$ and parameter space
$\mathbb{Q}$ and $\delta \Pi(\state) = 0$ is the force balance
equation given by \ref{}.

In order to apply optimisation algorithm...
We reduce the objective functional to be a function of the control
parameters only, $\hat{I}(\mvec) := I(\state(\mvec), \mvec)$. Consider an
initial value of the parmameters $\mvec = \mvec^0$. If $\hat{I}(m)$ is defined
and differentible in a neighborhood of $\mvec^0$ then $\hat{I}$ (and
consequently $I$) decreases fastest in the direction of the gradient, 
\begin{align}
  \nabla_{\mvec} \hat{I}(\mvec^0) = \begin{bmatrix}
    \frac{\mathrm{d} \hat{I}(\mvec^0) }{\mathrm{d} m_1},
    \frac{\mathrm{d} \hat{I}(\mvec^0) }{\mathrm{d} m_2},
    \cdots
    \frac{\mathrm{d} \hat{I}(\mvec^0) }{\mathrm{d} m_N}
  \end{bmatrix}^T.
  \label{eq:functional_gradient}
\end{align}
Here $\frac{\mathrm{d} \hat{I} }{\mathrm{d} m_i}$ represents the
total derivative:
\begin{align}
  \frac{\mathrm{d} \hat{I} }{\mathrm{d} m_i} = \frac{\mathrm{d} I (\state(\mvec), \mvec))}{\mathrm{d} m_i} = \frac{\partial  I }{\partial \state} \frac{\mathrm{d} \state}{\mathrm{d} m_i} + \frac{\partial  I }{\partial m_i}.
  \label{eq:functional_derivative_component}
\end{align}
In other words, the sequence
\begin{align}
  \mvec^{k+1} = \mvec^{k} - \gamma_k \nabla_{\mvec} \hat{I}(\mvec^k), \gamma_n \in \mathbb{R}
\end{align}
satisfies $\hat{I}(\mvec^{k+1}) \leq \hat{I}(\mvec^k) \; \forall k \geq
0$, and converges towards a local minimum. If $\hat{I}$ is convex then
the minimum is also global.
Being able to compute the gradient of the objective functional wrt to
the control parameters allows us to employ gradient based optimization
methods which are in general superior to gradient free methods.
One way to compute the gradient is by means of the finite difference
approach: For a given parameter $\mvec = \mvec^*$ we have
\begin{align}
  \frac{\mathrm{d} \hat{I} }{\mathrm{d} m_i}( \mvec^*) =
  \lim_{h \mapsto 0} \frac{\hat{I}(\mvec^* + h\mathbf{e}_i) - \hat{I}(\mvec^*)}{h}, 
\end{align}
where $\mathbf{e}_i \in \mathbb{Q}$ is the $i$'th canonical basis
vector. If $\dim(\mathbb{Q}) = N$, this approach would require $N+1$
functional evaluations. Moreover, since the state-variables depends upon the
control variables, we would also need to solve the force balance
equation $N+1$ times, which is typically very copmutationally
expensive. Hence, this approach is typically infeasable when the
dimesion of your parameterspace is large. In this case the adjoint
approach is much better. If $A$ is an operator (e.g a matrix), then
the adjoint operator $B$ satisfies the relation $\langle Au, v \rangle
= \langle u, Bv \rangle$, and we write $ B = A^*$. Here $(\cdot)^*$
denotes the Hermitian transpose, which in the case where $A$ is a real
matrix is just the transpose of $A$, $A^T$.

@@

>>>>

\section{Personalization of Cardiac Biomechanics}

In order to make use of models

\subsection{Data aquistition}

Since the discovery of X-ray in 1895, medical imaging has played a central
role in diagnosis, treatment planning and follow-up. Without the huge
advancements in medical imaging the last decades, computational models
would not have been as important and promising as it has become.

The interaction between cardiac images and cardiac models are important.

Still, one of the major bottlenecks in terms of personalized
computaional models is the quality of data (and representations of
model). There is a big issue in
reproducebility, because of noisy data, and operator depedent
measurements.

Today there are three main non-invasive imaging techniques used in
cardiology which is echocardiography (ECHO), magnetic
resonance imaging (MRI) and computed tomography (CT), and all of them
are used for aquiring data used in computaional modeling. Each
modality offers advantages and disadvatages over the other, and we
will make a short summary of pros and cons for each modelity below.

\subsubsection{CT}
Computed tomography is reconginzed as the

\subsubsection{ECHO}

\subsubsection{MRI}

For example, MRI provides
high quality images, uses zero radiation, but is exspensive and lacks
temporal resolution. CT can more accuratley reconstruct the 3D image
in contrast to MRI, in which 2D slices needs to be glued together to
form a 3D surface. However, CT exposes the patient to radiation which
increases the chance of developing cancer. Finally echocardiography is
easy to use, cheap, harmless, and  has good temporal resolution, but
is clearly inferior when it comes to image quality.

The main modality used in this thesis is 4D echocardiography, and we
will therefore focus on data aquired using this modality.
With 4D we mean three spatial and one temporal dimesion.
In order to aquire a 3D volume image using ultrasound, the speed of sound in
human tissue (which is approxmately 1540 m/s) put some limitations on
the image quality verus the frame-rate \cite{rabben2010technical}.
Consequently, a 3D ultracound image is formed by stiching together $N$
disjoint subvolumes aquired during $N$ cardiac cycles ($N$ typicall
between 2 and 8)\cite{brekke2007volume}.

The images are later processed using some image segmentation tool.
One such tool is EchoPac which is used for analysing images for GE
Vingmed. The 4D Auto LVQ tool in EchoPac, is a tool for processing 3D
echo images, and traces of volume, triangulated surfaces and 3D strain
traces together with a structured mesh of the AHA segments for each
time point, can be exported to and HDF format and later used for mesh generation.

@@

\subsubsection{Mesh generation}
In this section we will explain how to generate a left ventricular
mesh based on segmented surfaces coming from 4D echocardiography.
Figure \ref{fig:echopac_output} shows an example of how the exported data
from the image semgentation tool looks like. For each time point we
can extract triangulated surfaces of the endocardium and epicardium
which can be seen in Figure \ref{fig:echopac_out_surf}. Together with
these surfaces we are also given a so called strain mesh (Figure
\ref{fig:echopac_out_strain_mesh}) located approximately in the
midwall, which defines the approximate location of the AHA-segments
\cite{cerqueira2002standardized}, and can be used to orient the mesh.

\begin{figure}[htbp]
  \centering
  \begin{subfigure}[t]{0.45\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/geometry/raw.png}
    \caption{\label{fig:echopac_out_surf}}
  \end{subfigure}
  \begin{subfigure}[t]{0.45\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/geometry/strain_mesh.png}
    \caption{\label{fig:echopac_out_strain_mesh}}
  \end{subfigure}
\caption{}
\label{fig:echopac_output}
\end{figure}

As discussed in \ref{sec:mech_boudary}, we constrain the basal movement in
the longitudinal direction, which is easiest to accomplished using a flat
base located at a prescribed location. In order to make the base flat
we first orient the surfaces so that the longitudinal axis is aligned
with the $x$-axis, and the apex pointing in the postive $x$
direction. To conststruct the basal plane we first take out the basal
points from the strain mesh, and use these points to construct a least
sqauare fitting plane (Figure \ref{fig:strain_mesh_plane}). Let $(x_i, y_i, z_i), i = 1, \cdots, N$
be the basal points, and suppose the basal plane solves the equation
$z = ax + by + c$, for some unknown constants $a,b,c$. Following a
least square approach, we select the parameters $(a,b,c)$ that
minimizes to sum
\begin{align}
  \sum_{i = 1}^{N} \left( z_i - ax_i + by_i + c \right)^2.
\end{align}
Once the parameters for the basal plane is found we adjust the size of
the cut, by moving the plane along the longitudinal axis (here
$x-$axis), until the cavity volume agrees with the measured volume
given by the image segmentation tool within some specified tolerance. 
When the correct size is found the points above the plane is removed,
and the remaining surfaces are smoothed using the GAMer
\cite{yu2008feature}. The actualy mesh generation is performed using Gmsh
\cite{geuzaine2009gmsh}, which meshes the endocardial and epicardial
surface togther unsing frontal-Delaunay meshing algorithm. Gmsh also marks
the endocardal, the epicardial and the basal facets so that it is easy
to

\begin{remark}
  The cavity volume is computed by generating the whole mesh, and the
  compute the volume using Equation \eqref{eq:volume_operator}. The cut size is
  found by using a one dimensional optimization algorithm with
  the objective funcitonal representing the squared error between
  computed and measured volume.
\end{remark}

\begin{figure}[htbp]
  \centering
  \begin{subfigure}[t]{0.4\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/geometry/strain_mesh_plane.pdf}
    \caption{\label{fig:strain_mesh_plane}}
  \end{subfigure}
  \begin{subfigure}[t]{0.45\textwidth}
    \includegraphics[width=\textwidth]{chapters/introduction/figures/geometry/cut.png}
    \caption{\label{fig:cut}}
  \end{subfigure}
\caption{}
\label{fig:echopac_output}
\end{figure}

@@

\subsection{Data Assimilation}
The constitutive laws in cardiac mechanics comes from experiments
done on tissue slabs, and relates stresses in the material to strain.
If we had the same data available for a given patient we could easily
have performed a regression analysis in order to identify the
parameters in the contitutive model. However, this type of data
typically requires that tissue samples are taken from the myocardium,
which is not an option when dealing with living humans. In general,
when dealing with physical system such as the heart or weather
forecast it is not allways possible to design an experiment which
allows to determine all properties of a physical system
\cite{chapelle2013fundamental}. In such cases we have to take the
measurement that we have, and do our best to incorporate them into the
model.

For example, the type of data that we usually have available, are data
that can be extracted using imaging techniques such as motion data,
volume traces, regional strain traces etc. One technique to estimate
model parameters based on such observations is called data assimilation.

The field of data assimilation has its roots in meteorology, where
a typical problem is to make predictions on the weather, based on
observations (of temperature, humidity etc.) at different locations.

There are basically two main approaches to assimilate data;
\emph{sequential data assimilation} and \emph{variational data
  assimilation}. Sequential data assimilation, which is often
referred to as filtering, is based on a predictor-corector approach
where you start from your initial data and use a filter to predict the
next state. Once you have your measurement available you correct the
estimate based on the new observations. Some examples of sequential
procedures include Kalman filtering and Luenberger observers
\cite{chapelle2013fundamental}.

The other approach called variational data assimilation, which is the
one used in this thesis, is based on minimizing a cost functional that
represents the mismatch between observations and simulations, with
possibly additional regularization terms added to the objective
functional.

% The general setup for a variational data assimilation problem is the
% following. 
We will now explain the general setup for variational data
assimilation. In order to keep consistent notation we will refer to
the \emph{state variable} as $\mathbf{w}$, which is our case
represents jointly the displacement and hydrostatic pressure
$\mathbf{w}=(\mathbf{u}, p)$.

The state varibles are described by some \emph{phyiscal
model} $\mathcal{L}$, which is our case is governed by the force balance
equations \eqref{eq:force_balance}, and can be written in the form
$\mathcal{L}\mathbf{w} = f$.

We are also given a set of measurements (or observations) that we want
to assimilate, and we denote this by the vector $\mathbf{y}$.
For example, $\mathbf{y}$ might be a set of volume measurements or a
combinations of volume and strain at given time points in the cardiac
cycle. The \emph{observation operator} $\mathcal{H}$, is approximation
of the observation and act as an operator from the state space to the
observation space. We therefore have the relation
\begin{align}
  \mathbf{y} = \mathcal{H}(\mathbf{w}) + \xi_o, 
\end{align}
where $\xi_o$ represents both the error in the measurements (e.g
noise in the data) and error in the representation of the observation
(that $\mathcal{H}$ do not represent the data good enough).

\paragraph{The volume observation operator
  $\mathcal{H}_{\mathrm{volume}}$}
In some cases, the ventricular cavity volume can easily be computed
analytically from the displacement field. Denote the inner cavity of
the left ventricle in the current configuration by
$\omega_{\mathrm{endo}}$. Further let $\mathrm{d} v$ and $\xvec$
denote an infinitesimal volume element and the coordinate in the
current configuration respectively. Then by the divergece theorem we have
\begin{align}
  \mathcal{H}_{\mathrm{volume}}(\uvec)= \int_{\omega_{\mathrm{endo}}} \mathrm{d} v =
  \frac{1}{3}\int_{\omega_{\mathrm{endo}}} \nabla \cdot \xvec \mathrm{d} v =
  \frac{1}{3}\int_{\partial \omega_{\mathrm{endo}}} \xvec \cdot \mathbf{n} \mathrm{d} s, 
\end{align}
where $\mathbf{n}$ and $\mathrm{d} s$ are the unit normal and a surface
element on the current configuration respectively.
Note that the boundary $\partial \omega_{\mathrm{endo}}$ includes the
endocardial basal boundary. However in the case when the base is flat
and fixed in longitudinal direction, the contribution from this
boundary integral will be zero. In this case the boundary integral
will be the same as the integral over the endocardial boundary
decribed in \ref{refer_to_section}, with the only change being the
change of sign on the normal vector. By Nansens formula we obtain
\begin{align}
  \mathcal{H}_{\mathrm{volume}}(\uvec)= -\frac{1}{3}\int_{\lvendo} \left( \Xvec + \uvec \right) J \F^{-T} \Nvec \mathrm{d}S,
  \label{eq:volume_operator}
\end{align}
where now $\lvendo$,  $\Xvec$, $\Nvec$ and $\mathrm{d}S$ are respectively the
endocardial surface, the referece coordinate, the unit normal and a
surface element in the reference configuration.

\paragraph{The strain observation operator
  $\mathcal{H}_{\mathrm{volume}}$}
Another observation that is encountered in this thesis is strain
measurements, of more precisely average regional strain in the
circumferential, radial or longitudinal direction. Let $\Omega_j$
denote the volume for which the strain should be averaged over, and
let $\mathbf{e}_k$ denote the unit vectorfield in the preferred strain
direction. Then for a given strain tensor $\mathbf{A}$ we define 
\begin{align}
  \mathcal{H}_{\mathrm{strain}}(\uvec) = \frac{1}{|\Omega_j|}\int_{\Omega_j} \mathbf{e}_k^T \mathbf{A}(\uvec) \mathbf{e}_k  \mathrm{d}V
\end{align}
The strain tensor $ \mathbf{A}$ is typically chosen to be the
Grenn-Lagrange strain tensor $\mathbf{E}$\eqref{eq:green_lagrange}, or
the material displacement gradient tensor $\F - \I$.

\begin{remark}
  Measured strains are computed relative to some reference
  geometry, which do not always coinside with the reference geometry
  chosen for your simulation. In these cases you have to either
  recompute the measured strain according to the chosen referece for
  the simulation, or recompute the simulated strains according to the
  correct referece geometry for your measuments. 
\end{remark}

% Within varitaional data assimilation it is common to separate between
% dynamic models and static models. These are conventionally reffered to as
% $4D-$Var and $3D-$Var methods respectively. In the work done this
% thesis we have used both of these methods.

% For an introduction to data assimilation we refer to e.g \cite{lahoz2010data}.

Let us assume that

\begin{itemize}
  \item Examples of data you want to constrain
  \item Forming of mismatch functional and problem formulation as
    PDE-constrained optimization.
  \item The adjoint approach. comparison with other methods.
  \item Control parameters
  \item OPtimization algorithm - stopping criteria
  \item Reference to pulse{\_}adjoint and pulse{\_}adjoint{\_}post
    repositories
  \item Variational versus sequnetial
  \item Question about uniquness
\end{itemize}

@@

\begin{equation}
  \begin{aligned}
    \label{eq:opt_matparam}
    & \underset{\mvec}{\text{minimize}}
    & &  \mathcal{J}(\state, \mvec) \\
    & \text{subject to}
    & & \delta \Pi(\state, \mvec) = 0, 
  \end{aligned}
\end{equation}

where $\mathcal{J}(\state, \mvec): \mathbb{W} \times \mathbb{Q} \mapsto
\mathbb{R}$ for some state space $\mathbb{W}$ and parameter space
$\mathbb{Q}$ and $\delta \Pi(\state, \mvec)) = 0$ is the force balance
equation given by \ref{}.
$\mvec = (m_1, \cdots m_P)$

Now suppose we have discretized the partial differental equation
$\delta \Pi(\state, \mvec) = 0$, which reduces to a system of the form
$\mathbf{A} \state = \mathbf{b}$, with possible all terms depending on
the parameters $\mvec$. The gradient of the cost functional with
respect to the parameter is given by
\begin{align}
  \frac{\mathrm{d} \mathcal{J}}{\mathrm{d} \mvec} = \frac{\partial \mathcal{J}}{\partial \mvec}
  + \frac{\partial \mathcal{J}}{\partial \state} \frac{\mathrm{d} \state}{\mathrm{d} \mvec}.
  \label{eq:grad_cost}
\end{align}
Here $\frac{\mathrm{d}}{\mathrm{d} \mvec}$ denotes the total
derivative, while $ \frac{\partial}{\partial \mvec}$ denotes the
partial derivative. Computing the partial derivatives in
\eqref{eq:grad_cost} are usually straight forward, while computing  $
\frac{\mathrm{d} \state}{\mathrm{d} \mvec}$ is hard. To see this, let
us differentiate the equation $\mathbf{A} \state = \mathbf{b}$ with
respect to $m_i$, and solve for the

\begin{remark}
  In data assimilation we typically want to copmute the gradient of
  the cost function because we want to employ gradient-based
  optimization algorithms for finding the ``optimal'' parameter
  set. In other cases, the gradient might be important because it
  tells you something about the sensitivity of your function of
  interest with respect to the parameters. For example, if you have a
  model with many input parameters, you may want to identify what
  impact the different parameters has on the output. Perturbing each
  parameter and observing the chane in the objective function quickly
  becomes infeasible when the number of parameters becomes large, in
  which case the adjoint aproach offer a possible solution. 
\end{remark}

\subsubsection{Adjoint method}
In order to apply optimisation algorithm...
We reduce the objective functional to be a function of the control
parameters only, $\hat{I}(\mvec) := I(\state(\mvec), \mvec)$. Consider an
initial value of the parmameters $\mvec = \mvec^0$. If $\hat{I}(m)$ is defined
and differentible in a neighborhood of $\mvec^0$ then $\hat{I}$ (and
consequently $I$) decreases fastest in the direction of the gradient, 
\begin{align}
  \nabla_{\mvec} \hat{I}(\mvec^0) = \begin{bmatrix}
    \frac{\mathrm{d} \hat{I}(\mvec^0) }{\mathrm{d} m_1},
    \frac{\mathrm{d} \hat{I}(\mvec^0) }{\mathrm{d} m_2},
    \cdots
    \frac{\mathrm{d} \hat{I}(\mvec^0) }{\mathrm{d} m_N}
  \end{bmatrix}^T.
  \label{eq:functional_gradient}
\end{align}
Here $\frac{\mathrm{d} \hat{I} }{\mathrm{d} m_i}$ represents the
total derivative:
\begin{align}
  \frac{\mathrm{d} \hat{I} }{\mathrm{d} m_i} = \frac{\mathrm{d} I (\state(\mvec), \mvec))}{\mathrm{d} m_i} = \frac{\partial  I }{\partial \state} \frac{\mathrm{d} \state}{\mathrm{d} m_i} + \frac{\partial  I }{\partial m_i}.
  \label{eq:functional_derivative_component}
\end{align}
In other words, the sequence
\begin{align}
  \mvec^{k+1} = \mvec^{k} - \gamma_k \nabla_{\mvec} \hat{I}(\mvec^k), \gamma_n \in \mathbb{R}
\end{align}
satisfies $\hat{I}(\mvec^{k+1}) \leq \hat{I}(\mvec^k) \; \forall k \geq
0$, and converges towards a local minimum. If $\hat{I}$ is convex then
the minimum is also global.
Being able to compute the gradient of the objective functional wrt to
the control parameters allows us to employ gradient based optimization
methods which are in general superior to gradient free methods.
One way to compute the gradient is by means of the finite difference
approach: For a given parameter $\mvec = \mvec^*$ we have
\begin{align}
  \frac{\mathrm{d} \hat{I} }{\mathrm{d} m_i}( \mvec^*) =
  \lim_{h \mapsto 0} \frac{\hat{I}(\mvec^* + h\mathbf{e}_i) - \hat{I}(\mvec^*)}{h}, 
\end{align}
where $\mathbf{e}_i \in \mathbb{Q}$ is the $i$'th canonical basis
vector. If $\dim(\mathbb{Q}) = N$, this approach would require $N+1$
functional evaluations. Moreover, since the state-variables depends upon the
control variables, we would also need to solve the force balance
equation $N+1$ times, which is typically very copmutationally
expensive. Hence, this approach is typically infeasable when the
dimesion of your parameterspace is large. In this case the adjoint
approach is much better. If $A$ is an operator (e.g a matrix), then
the adjoint operator $B$ satisfies the relation $\langle Au, v \rangle
= \langle u, Bv \rangle$, and we write $ B = A^*$. Here $(\cdot)^*$
denotes the Hermitian transpose, which in the case where $A$ is a real
matrix is just the transpose of $A$, $A^T$.

@@

\subsubsection{Dolfin-Adjoint}

Dolfin-Adjoint is a software package that based on the FEniCS project
which aims to derrive the discrete adjoint equation and the tangent
linear models of finite element models implemented within the FEniCS
framework. While the tradional approach is to derrive the adjoint code
from the forward code, Dolfin-Adjoint utilises the high level symbolic
representation \cite{UFL} in FEniCS to derrive the discrete adjoint equations
from the discrete forward eqations, and then the FEniCS system
derrives the adjoint code from the discrete adjoint equations.

\subsubsection{Optimization methods}

\subsection{Identifiability of parameters}
Try dividing heart mesh is different segment, discuss uniqueness of
paraperers
Discuss regularization technices

\subsection{Mehcanical biomarkers}
A word about what models can contribute with

### 2017-03-24 chapters/introduction/physiology.tex
Update thesis to latest version

The heart is the muscular organ responsible for circulating the blood
thoughout the body.

Deoxgenated blood transported from the body trough the veins to the
right atrium (RA)

@@

@@

>>>>

The heart is the muscular organ responsible for circulating the blood
throughout the body.

Deoxgenated blood is transported from the body trough the veins and to
the right atrium (RA). From the RA, blood flows to the right ventricle
(RV), before it is sent through the the pulmonary ateries(PA) and to the
lungs, where the bloods gets oxygenated. From the lungs, the blood
travels trough the pulmonary veins to the left atrium (LA), and
finally to the left ventricle (LV), before is is sent out to the body
again trough the aorta (Ao). A schamtic view of the blood flow
through the body is shown in Figure ~\ref{}

The flow from one chamber to another is driven
by pressure gradients, that is blood flows from regions with higher
pressure to regions with lower pressure. Consequently, the lowest
pressure is found in the veins entering the heart, and the highest
pressure is found in the left ventricle.

The heart itself is located approximately in the center of your chest
with the \emph{apex} of heart angled down towards the left of the
body, and the \emph{base} located just behind the breastbone.

@@

Starting form the outside of the heart we have the \emph{pericardium},
the \emph{epicardium}, the \emph{myocardium} and the \emph{endocardium}. 
The endo- and epicardium are respectively the inner and outer layers of the
myocardium, which is the cardiac muscle cells.

The myocardium is made up by individual cardiac muscle cells, or
cardiac myocytes which branch and join neighbouring cell, and form a
complicated fibrous network which is often referred to as a
\emph{functional syncytium}; the number of active muscle fibers cannot
be varied from beat to beat.

Each myocyte is composed of bundles of myofibriles, which again
contain \emph{myofilaments}. The myofilaments cosists of a repeated pattern
of lines and bands which can be seen as a collection of individual
contrictile units called \emph{sarcomeres}.

\subsection{Cardiac pump function}

\subsection{Frank-Starling Mechanism and the Starling curve} 
The end diastolic pressure volume relationship (EDPVR), relates the end
diastolic volume to the end diastolic pressure. Simlarly the end
systolic pressure volume relationship (ESPVR) relates the end systolic
volume to the end systolic pressure. The Frank-Starling relationship,
relates the EDPVR to the ESPVR. The law states that the stroke volume
increases in response to and increase in end diastolic volume (as long
as the sarcomeres are working on the acending limb, which is usually
the case). This allows the cardiac output to be synchronized with the
venous return and blood supply without depending upon external
regulation.
The Starling Curve is a plot of the ESPVR - EDPVR, showing that the
output increases up to some maximum (at the acending limb). Thereafter
it decreases on the decending limb. If the sarcomeres works on the
decending limb, the heart would dilate and become weakend.

@@

Contractility can be seen as the manifestation of all the factors that
influence the interactions between the contractile proteins.
One simple definitions is \emph{myocardial sontractility is the
  ability of the heart to do work}. A change in myocardial
contractility can be seen if there is a change in ejection which is
not caused by changes in the initial fiber lenght (ED). If the end
diastolic volume remains the same, but the stroke work changes there
is a change in contractility.

% Under normal circumstances, the myocardial cells generate less than
% half the of their maximal contractility. One could increase the
% contractility by providing more calcium to the cells.

In order to find a suitable idex for myocardial contractility we take
step back and consider Hills postulate: \emph{the ``active points'' of
  muscle can exist in two states: 1. active points are maintaining
  tension, 2. the active points are participating in a chemical
  reaction.} 
If the muscle is presented with a heavy load, developed tesion is
high, shortening velocity is low and the load is only moved a short
distance. In this case the active points are holding tesion.
On the other hand, if the muscle is presented with a light load, the
shortening velocity is high and the load is moved a longer
distance. In this case more of the active points are participating in
chemical reactions. 
In both cases, the contractility might be the same which is why we
need a load independent measure of contractility.
Looking at the force-velocity relationship we can say that the maximum
force (tesion) is acheived when all the active points are holding
tesion ($P_0$). $P_0$ is proportional to the number of active
cross bridges.
Similarly, the maximum velocity is acheived when all
the active points are participating in a chemical reaction
($V_{\text{max}}$). $V_{\text{max}}$ reflects the maximal rate of
cross bridge cycling. 
This is explained by the Hill eqation, which
relates the force to velocity, and the interects are $P = P_0, (V = 0)$
and $V = V_{\text{max}}, (P = 0)$. This is very difficult to meause in
cardiac cells because cariac muscle cells are not tentanized.

The most commonly used index of contractility was formalized
\cite{sagawa1977end} using a time varying elastance model, which
states that at any point in the cardiac cycle the pressure and volume
are linearly related by the time-varying elastance, or formally:
\begin{align}
  P(t) = E(t)( V(t) - V_0 ).
  \label{eq:time_varying_elastance}
\end{align}
Here $P$ is the endocardial pressure, $V$ the cavity volume, $E$ the
time-varying elastance, and $V_0$ the volume axis intercept. This
model
